Perturbative theories in the complex plane

### 2020-06-28 SlideToulouse/main.tex
Add files via upload

%% General document %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{decorations.fractals}
\usepackage{mathpazo}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{xcolor}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage{physics}
\usepackage{multimedia}
\usepackage{subfigure}
\usepackage[absolute,overlay]{textpos}
\usepackage{ragged2e}
\usepackage{amssymb}
\usepackage[version=4]{mhchem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title[Title]{Perturbation theories in the complex plane}
\author[]{Antoine \textsc{Marie}}
  \setbeamersize{text margin left=5mm}
  \setbeamersize{text margin right=5mm}
\institute{Supervised by Pierre-François \textsc{LOOS}}

@@

\date{30th June 2020}
\titlepage
\end{frame}

@@

\begin{frame}{Why do we use perturbation theories in computational chemistry?}

\pause[1]

The Hartree-Fock theory is \textcolor{Green4}{computationally cheap} and can be applied even to \textcolor{Green4}{large systems}.

But this method is missing the \textcolor{red}{correlation energy}...

\pause[2]

$\rightarrow$ We need methods to get this correlation energy!

\pause[3]

\begin{beamerboxesrounded}[scheme=foncé]{\centering A general method}
In physics perturbation theory is often a good way to improve the obtained results with an approximated Hamiltonian.
\end{beamerboxesrounded}

@@

\begin{frame}{The Møller-Plesset perturbation theory}

\pause[1]

@@

\begin{itemize}
\centering
    \item $H_0$: Unperturbed Hamiltonian
    \item $V$: Perturbation operator
\end{itemize}

\pause[2]

@@

\begin{equation}
   F = T + J + K
\end{equation}

\begin{itemize}
\centering
    \item $T$: Kinetic energy operator
    \item $J$: Coulomb operator
    \item $K$: Exchange operator
\end{itemize}

\pause[3]

\begin{beamerboxesrounded}[scheme=foncé]{}
\centering
Full Configuration Interaction gives us access to high order terms of the perturbation series !
\end{beamerboxesrounded}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{gill1986.png}
    \caption{\centering Barriers to homolytic fission of \ce{He2^2+} using minimal basis set MPn theory (n~=~1-20).\footnote{\cite{gill_deceptive_1986}}}
    \label{fig:my_label}
\end{figure}

@@

\begin{frame}{Multi-reference and spin contamination}
\begin{table}
    \centering
    \begin{tabular}{c c c c c c c}
\hline
 $r$ & UHF & UMP2 & UMP3 & UMP4 & $\expval{S^2}$ \\
\hline
0.75 & 0.0\% & 63.8\% & 87.4\% & 95.9\% & 0.00\\
1.35 & 0.0\% & 15.2\% & 26.1\% & 34.9\% & 0.49\\
2.00 & 0.0\% & 01.0\% & 01.8\% & 02.6\% & 0.95\\
2.50 & 0.0\% & 00.1\% & 00.3\% & 00.4\% & 0.99\\
\hline
\end{tabular}
    \caption{\centering Percentage of electron correlation energy recovered and $\expval{S^2}$ for the \ce{H2} molecule as a function of bond length (r,\si{\angstrom}) in the minimal basis.}
    \label{tab:my_label}
\end{table}

@@

But the Taylor expansion of this function does not converge for $x\geq1$... 
\vspace{0.3cm}
\centering Why ?

@@

\begin{equation*}
   H(\lambda) = H_0 + \lambda V
\end{equation*}

@@

\begin{itemize}
    \item Partitioning of the Hamiltonian: Möller-Plesset, Epstein-Nesbet,...
    \item Zeroth order reference: weak correlation or strongly correlated electrons.
    \item Finite or complete basis set.
    \item Localized or delocalized basis functions.
\end{itemize}

@@

\begin{beamerboxesrounded}[scheme=foncé]{A 2x2 matrix}
\centering \small{$\mqty(\alpha & \delta \\ \delta & \beta)  =$}

\vspace{0.15cm}

\small{$\mqty(\alpha + \alpha_s & 0 \\ 0 & \beta + \beta_s ) + \mqty(- \alpha_s & \delta \\ \delta & - \beta_s)$}

@@

\begin{frame}{Two-state model}

@@

For $\lambda<0$:

\begin{equation*}
    H(\lambda)=\sum\limits_{j=1}^{2n}\left[ \underbrace{-\frac{1}{2}\grad_j^2 - \sum\limits_{k=1}^{N} \frac{Z_k}{|\vb{r}_j-\vb{R}_k|}}_{\text{Independant of }\lambda} + \overbrace{(1-\lambda)V_j^{(scf)}}^{\textcolor{red}{Repulsive}}+\underbrace{\lambda\sum\limits_{j<l}^{2n}\frac{1}{|\vb{r}_j-\vb{r}_l|}}_{\textcolor{blue}{Attractive}}  \right]
\end{equation*}

\footnote{stillinger, sergeev, baker}

@@

\pause[1]

\begin{beamerboxesrounded}[scheme=foncé]{\centering Exact energy $E(z)$}
$E(z)$ has a critical point on the negative real axis and $E(z)$ is continue for real values below $z_{crit}$.
\end{beamerboxesrounded}

@@

\pause[2]

@@

\pause[3]

\centering \Large{How is this connected???}

@@

\pause[1]

@@

\pause[2]

\begin{beamerboxesrounded}[scheme=foncé]{\centering Singularity $\alpha$}
\begin{itemize}
    \item Large avoided crossing
    \item Non-zero imaginary part
    \item Interaction with a low lying doubly excited states
\end{itemize} 
\end{beamerboxesrounded}

\pause[3]

\begin{beamerboxesrounded}[scheme=foncé]{\centering Singularity $\alpha$}
\begin{itemize}
    \item Sharp avoided crossing
    \item Very small imaginary part
    \item Interaction with a diffuse function 
\end{itemize} 
\end{beamerboxesrounded}

\footnote{sergeev}

@@

\pause[1]

\begin{beamerboxesrounded}[scheme=foncé]{\centering Stillinger}
\begin{quote}
    \textit{"One might expect that $E_{FCI}(z) $ would try to model a continuum at $z_c$ with a grouping of discrete but closely spaced eigenstates that undergo sharp avoided crossing with the ground states."}
\end{quote}
\end{beamerboxesrounded}

\pause[2]

\begin{beamerboxesrounded}[scheme=foncé]{\centering Sergeev et al.}
Proof of the existence of this group of sharp avoided crossings for Ne, He and HF when the basis set contains diffuse functions.
\end{beamerboxesrounded}

@@

\begin{beamerboxesrounded}[scheme=foncé]{\centering Two electrons on a sphere Hamiltonian}
\begin{equation*}
    H=-\frac{1}{2}(\grad_1^2 + \grad_2^2) + \frac{1}{r_{12}}
\end{equation*} 
\end{beamerboxesrounded}
\vspace{0.5cm}

@@

\begin{frame}{Apparition of a class $\beta$ singularity}

\pause[1]

\begin{beamerboxesrounded}[scheme=foncé]{\centering Expectation}
The electrons are restricted to the surface of the sphere so we should not observe singularities characteristic of ionization processes.
\end{beamerboxesrounded}

\pause[2]

\large But for some values of R... we actually observe some $\beta$ singularities!
\centering Why?

\pause[3]

\begin{beamerboxesrounded}[scheme=foncé]{\centering Symmetry breaking}
The $\beta$ singularities observed are connected to the symmetry breaking of the wave function.
\end{beamerboxesrounded}

@@

\pause[1]

\begin{beamerboxesrounded}[scheme=foncé]{\centering Møller-Plesset perturbation theory}
By understanding how the singularities are localized in the complex plane we hope that it will gives us a deep understanding of the strengths and weaknesses of the Møller-Plesset method to get the correlation energy.
\end{beamerboxesrounded}

\pause[2]

But there is an other secret application of exceptional points...

\pause[3]

\begin{beamerboxesrounded}[scheme=foncé]{\centering A new way to excited states energies}
The exceptionnal points connect ground and excited states in the complex plane. Using those properties one can smoothly morph a ground state in an excited state. 
\end{beamerboxesrounded}

>>>>

%% General document %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{decorations.fractals}
\usepackage{mathpazo}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage{physics}
\usepackage{multimedia}
\usepackage{subfigure}
\usepackage{xcolor}
\usepackage[absolute,overlay]{textpos}
\usepackage{ragged2e}
\usepackage{amssymb}
\usepackage[version=4]{mhchem}

\renewcommand{\thefootnote}{\alph{footnote}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title[Title]{Perturbative theories in the complex plane}
\author[]{Antoine \textsc{Marie}}
\date{30 Juin 2020}
  \setbeamersize{text margin left=5mm}
  \setbeamersize{text margin right=5mm}
\institute{Supervised by Pierre-François \textsc{LOOS}}

@@

\date{24 Avril 2020}
\titlepage
\end{frame}

@@

@@

\begin{frame}{The Möller-Plesset theory}

@@

\begin{itemize}
\centering
    \item $H_0$ : Unperturbed Hamiltonian
    \item $V$ : Perturbation operator
\end{itemize}

@@

\begin{equation}
   F = T + J + K
\end{equation}
\end{beamerboxesrounded}

\begin{itemize}
\centering
    \item $T$ : Kinetic energy operator
    \item $J$ : Coulomb operator
    \item $K$ : Exchange operator
\end{itemize}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{gill1986.png}
    \caption{\centering Barriers to homolytic fission of \ce{He2^2+} using minimal basis set MPn theory (n~=~1-20).}
    \label{fig:my_label}
\end{figure}

\footnotetext{\tiny{Gill et al.~Deceptive convergence in Møller-Plesset perturbation  energies, \textit{Chemical Physics Letter}, 1986}}

@@

\begin{frame}{Multi-reference and spin contamination}
\begin{table}
    \centering
    \begin{tabular}{c c c c c c c}
\hline
 $r$ & UHF & UMP2 & UMP3 & UMP4 & $<S^2>$ \\
\hline
0.75 & 0.0\% & 63.8\% & 87.4\% & 95.9\% & 0.00\\
1.35 & 0.0\% & 15.2\% & 26.1\% & 34.9\% & 0.49\\
2.00 & 0.0\% & 01.0\% & 01.8\% & 02.6\% & 0.95\\
2.50 & 0.0\% & 00.1\% & 00.3\% & 00.4\% & 0.99\\
\hline
\end{tabular}
    \caption{\centering Percentage of electron correlation energy recovered and $<S^2>$ for the \ce{H2} molecule as a function of bond length (r,A) in the minimal basis.}
    \label{tab:my_label}
\end{table}

@@

But the Taylor expansion of this function does not converge for $x\geq1$ ... 
\vspace{0.3cm}
\centering Why ?

@@

\begin{equation*}
   H = H_0 + \lambda V
\end{equation*}

@@

\begin{itemize}
    \item Partitioning of the Hamiltonian: Möller-Plesset, Epstein-Nesbet, ...
    \item Zeroth order reference: weak correlation or strongly correlated electrons.
    \item Finite or complete basis set.
    \item Localized or delocalized basis functions.
\end{itemize}

@@

\begin{beamerboxesrounded}[scheme=foncé]{A 2x2 matrix\textsuperscript{a}}

$

\small{\centering \begin{pmatrix}
   \alpha & \delta \\
   \delta & \beta 
\end{pmatrix} =

\vspace{0.3cm}

\begin{pmatrix}

   \alpha + \alpha_s & 0 \\
   0 & \beta + \beta_s
\end{pmatrix} +
\begin{pmatrix}
   - \alpha_s & \delta \\
   \delta & - \beta_s
\end{pmatrix}}
$

@@

\begin{frame}{Two state model}

@@

For $\lambda<0$ :

\begin{equation*}
    H(\lambda)=\sum\limits_{j=1}^{2n}\left[ \underbrace{-\frac{1}{2}\nabla_j^2 - \sum\limits_{k=1}^{N} \frac{Z_k}{|\vb{r}_j-\vb{R}_k|}}_{\text{Independant of }\lambda} + \overbrace{(1-\lambda)V_j^{(scf)}}^{\textcolor{red}{Repulsive}}+\underbrace{\lambda\sum\limits_{j<l}^{2n}\frac{1}{|\vb{r}_j-\vb{r}_l|}}_{\textcolor{blue}{Attractive}}  \right]
\end{equation*}

@@

\begin{beamerboxesrounded}[scheme=foncé]{\centering Exact energy $E(z)$}
$E(z)$ has a critical point on the negative real axis and $E(z)$ is continue for real value below $z_{crit}$.
\end{beamerboxesrounded}

@@

@@

\centering \Large{How is this connected ???}

@@

@@

\begin{beamerboxesrounded}[scheme=foncé]{\centering Singularity $\alpha$}
\begin{itemize}
    \item Large avoided crossing
    \item Interaction with a low lying doubly excited states
    \item Non-zero imaginary part
\end{itemize} 
\end{beamerboxesrounded}

\begin{beamerboxesrounded}[scheme=foncé]{\centering Singularity $\alpha$}
\begin{itemize}
    \item Sharp avoided crossing
    \item Interaction with a diffuse function 
    \item Very small imaginary part
\end{itemize} 
\end{beamerboxesrounded}

@@

@@

\begin{beamerboxesrounded}[scheme=foncé]{\centering Two electrons on a sphere Hamiltonian}
\begin{equation*}
    H=-\frac{1}{2}(\nabla_1^2 + \nabla_2^2) + \frac{1}{r_{12}}
\end{equation*} 
\end{beamerboxesrounded}
\vspace{0.5cm}

@@

\begin{frame}{Why is there a class $\beta$ singularity ?}

@@
